--- 
title: "Regression and ANOVA"
author: "Justin Patterson"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
citation-style: nature.csl
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a guide to ANOVA and regression and the connections between them."
---

# Prerequisites {-}

This project is meant to be a personal guide to ANOVA and regression. The scope of this project does not include time series analysis. Also, the focus will not be on designing experiments, but rather on analyzing the data from experiments which have already been conducted. To accomplish this, we will use simulated experimental data.

**Disclaimer**
This project was not written by an expert. That being said, I would appreciate any comments.

This book is written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
